% !TeX root = ../incremental_SS_Translation.tex

\chapter{Uttaratantra 64: Preservation of Health}% svastharakṣanīya

\section{Literature}

Meulenbeld offered an annotated overview of this chapter, which is called 
\emph{svasthavṛtta} in the vulgate.\fvolcite{IA}[330--331]{meul-hist} There, 
he translates \emph{svasthavṛtta} as ``preservation of health'' but elsewhere 
as ``maintenance of health''\fvolcite{IA}[301]{meul-hist} and 
``hygiene.''\fvolcite{IB}[738]{meul-hist} His only reference to secondary 
literature on this topic is to an annotated German translation by J. Laping 
1984.\footnote{\cite{lapi-1984}, cited by \volcite{IB}[431]{meul-hist}.} His 
footnotes to the chapter also indicate 
many parallels with \emph{Carakasaṃhitā, Sūtrasthāna}, chapter six.

On the term \emph{svastha}, Meulenbeld cites W. Halbfass (1991): 250-252; M. 
Hara (1995) and P.V. Sharma (1985e: 52-53).\fvolcite{IB}[10]{meul-hist}

In primary sources, see the definition of \emph{svasthā} in 
\emph{Suśrutasaṃhitā, Sūtrasthāna} 15.41--44; \emph{svasthahita} in 
\emph{Aṣṭāṅgahṛdaya, Sūtrasthāna} 1.16cd; ...

Comparable chapters in other medical texts include the \emph{svasthacatuṣka} 
in \emph{Carakasaṃhitā, Sūtrasthāna} (see below) and a chapter on 
\emph{svasthavṛtti} in Cakrapāṇidatta's \emph{Cakradatta}.

Meulenbeld also refers to \emph{svasthacatuṣka}, a group of four chapters 
(5--8) in the \emph{Carakasaṃhitā, Sūtrasthāna}, chapters 5–8, also called 
\emph{svasthavṛttacatuṣka} or \emph{svāsthyavṛttikacatuṣka} in various 
colophons of the text and commentaries. He summarises these 
chapters,\fvolcite{IA}[13--14]{meul-hist} and their names are 
\emph{mātrāśitīya}, daily regime (cf. Bhela Sū.8; A.h.Sū.8; A.s.Sū.11. Cf. 
Su.Ci.24); \emph{tasyāśitīya}, seasonal regimen (\emph{ṛtucaryā}) 
(Meulenbeld cites many references in primary and secondary 
literature\fvolcite{IB}[13]{meul-hist}); \emph{navegāndhāraṇīya}, the 
unwholesome effects of suppressing natural urges, etc. (Cf. Bhela Sū.6; 
Su.Ci.24); and \emph{indriyopakramaṇīya}, senses, mind and rules of conduct 
(Cf. Bhela Sū.7; Su.Ci.24. See on Ca.Sū.8: S.Ch. Vidyabhusana (1971): 11--12).

Meulenbeld mentions that a ``Sadānanda wrote a commentary, called \emph{Auṣadhavivṛti}, on the chapters of the
\emph{Carakasaṃhitā} containing rules for the preservation of health 
(svasthavṛtta).\fvolcite{IA}[196]{meul-hist} His footnote on this work refers to 
the NCC (VI, 396–397) and the following bibliographic reference: 
\emph{pañcatantram \ldots\  
carakasūtrasthānasvasthavṛttacatuṣkākhyacaturadhyāyāḥ\ldots\
sadānandaśāstrikṛtauṣadhavivṛtiyutayā saṃvalitaṃ}, Mercantile Press, Lahore 
1926 [IO.San.D.554].\fvolcite{IB}[302]{meul-hist}

In the vulgate, this chapter is included in a section called, ``the chapters that 
are an embellishment of the text" (\emph{tantrabhūṣaṇādhyāya}). However, in 
the Nepalese version, it is the eighth part of tenfold section on treatment for the 
body (\emph{kāyacikitsita}). See K, folio 209r, ll. 4--5 
(\emph{mūtradoṣo mūtrāghāto yonyamānūṣa eva ca/  
apasmāronmādakañ caiva rasabhedas tathaiva ca// 
svastharakṣāvidhāṇañ ca tantrayuktiś ca doṣabhit/
ity ebhir daśakaṃ proktaṃ punaḥ kāyacikitsite//}).


\section{Translation}

\begin{translation}    
  
\item [1] 
Now we shall explain how to preserve a person's health.%
% the upkeep of a healthy person.% rakṣaṇīya - maintenance of? prophylaxis for a healthy person?
	\footnote{The term \emph{svastharakṣaṇīya} more literally means, ``preserve a healthy person.'' \Dalhana{6.64.1}{808} noted this reading, which is also what we see in the Nepalese version, and explained it in these terms.
    
    The word \emph{rakṣaṇīya} (preservation, maintenance, upkeep, etc.) is replaced in the vulgate \citep[808]{vulgate} by \emph{vṛtta}, which, when compounded with \emph{svastha}, denotes the “regime for a healthy person.” However, \emph{rakṣaṇa} is retained later in the same chapter of the vulgate.} 

\item [2] 
One whose humours, digestive fire, and the functioning of bodily constituents and waste products are in equilibrium; whose self, senses, and mind are serene is called healthy.\footnote{This verse is absent here in the vulgate but occurs at the end of \emph{Sūtrasthāna} 15 in both the Nepalese version and vulgate. The vulgate \citep[808]{vulgate} supplements its omission with the statement, ``Such a healthy person has been described in the \emph{Sūtrasthāna}'' (\emph{sūtrasthāne samuddiṣṭaḥ svastho bhavati yādṛśaḥ}). \Dalhana{1.15.41}{75--76} discussed several points of view
regarding the meaning of equilibrium in these different 
contexts.}\q{More on Ḍalhaṇa and the meaning of sama?}

\item [3] 
My dear, the preservation of a healthy person is the supreme benefit of medicine. The topic of his regime and preservation, which was mentioned by me at the beginning, was outlined in brief and will here be explained in detail.
% who are the interlocutors? Divodāsa and Suśruta? See Chapter 66
% something has gone wrong with artham in the Nepalese manuscripts (which has arthāṃ and arthān). The participle (uktaṃ) and verb (vakṣyate) require a singluar. And artham can be neuter. The syntax of the vulgate is wrong because the subject is plural (arthāḥ) and the verb singular (a patch made inadequate by metric causa?)
% ḍalhaṇa: āditaḥ anāgatabādhe

\item [4] 
In each season, when specific humours are aggravated in people, a wise physician should prescribe the appropriate tastes for that season.

\item [5] 
You should know that when people's digestive fire is weak because of the dampness of their bodies during the rainy months, wind and other humours become aggravated due to their excitement.
% Ḍalhaṇa appears to understand mārutādayaḥ as referring to the doṣas.
% have I over-translated khalu?

\item [6--8ab]
Therefore, in order to remove moisture and maintain the humours, the king should be given solid food with astringent, bitter, and pungent tastes, which is neither too oily nor excessively dry, and which is heating and stimulates digestive fire. He should drink the water which was mentioned before,%
        \footnote{Ḍalhaṇa\citep[809]{vulgate} understands “the water which was mentioned” as monsoon rains, which was described in the section on what to drink (\emph{yaj jalaṃ coktam ādita iti ‘varṣāsv antarikṣaṃ’ ityādinā pānīyavarge kathitam}).} %
or water that has been heated by hot iron and mixed with honey. 
% I am not sure about the yat in the fourth pāda

\item [8cd--9]
On a rainy and windy day that is affected by excessively cold water,  herbal medicines become acidic if they are fresh.%
	\footnote{Ḍalhaṇa \citep[809]{vulgate}  glosses \emph{vidāha} (“burning”) as \emph{amlapāka} (“sour digestion”), suggesting that fresh herbs may cause acidic irritation in the stomach when taken on a rainy, wind and cold wet day.}
	%
And because of that, a wise person should not undertake excessive physical exercise.

\item [10] % not in the vulgate
When it is not very cold, aged barley, wheat and rice with pungent broths and soups, as well as appetising wild meats, should be eaten. 
% emended to bhojyā based on Caraka Sū 6.38c (note in the edition)
% not sure of the standard translation of kaṭu as a taste (bitter, sharp, pungent?). 
% the ca in pāda 4 suggests that jāṅgala is wild meat/deer/game rather than an adjective. Elsewhere in Ḍalhaṇa I see eṇādayo jāṅgalamṛgamāṃseṣu (an adjective here but obviously refers to game).

\item [11] % not in the vulgate
One should drink a little mead or sweet wine, %
	\footnote{On \emph{mādhvīka}, a kind of honey liquour, see McHugh 2021: 46. On \emph{sīdhu} (“sweet wine”), Ḍalhaṇa \citep[778]{vulgate} says it is \emph{gauḍika}, an alcoholic drink produced from sugar or grain (\emph{gauḍikaṃ sīdhu guḍakṛtaṃ sasyakam}). McHugh (2021: 45–46), who translates \emph{sīdhu} as “sugarcane wine,” says that it is primarily made from sugarcane but also other sweet substances, like jaggery, herbs and mahua flowers.}
%
and also rainwater, lake water, or well water, hot or cold and mixed with honey. 

\item[12] % ab not in the vulgate
In this climate (\emph{atra}), one should avoid the easterly wind, exposure to the heat, buttered barley water, river water, sleep during the day, frost, and sex.

\item[13--14ab] % not in the vulgate
One should engage in body rubs, massages, bathing, perfumes, garlands and smoking. To avoid humidity from the ground, one should have a heavy cloak and sleep in a house's central room, which is cool, has a fire, and is free from draughts.
%
% uṣma heat, hot vapour, steam. Is "humidity" a better translation for uṣma  here as I think this still concerns the varṣa season? I now see that PV Sharma translates it as "ground humidity." Cf, alternative verse in the vācaspatyam: bhūvāṣpaparihārārthaṃ śayīta ca vihāyasi.
% cf: Caraka1.6.40: pragharṣodvartanasnānagandhamālyaparo bhavet | laghuśuddhāmbaraḥ sthānaṃ bhajed akledi vārṣikam || [during the rainy season]
% cf: AS.Sū.4.48	pragharṣodvartanasnānadhūmagandhāgarupriyaḥ | 
%				yāyāt kareṇumukhyābhiścitrasragvastrabhūṣitaḥ || [varṣa season]
% Jayaratha on Tantrāloka: madhyegṛhamiti--gṛhamadhye brahmasthāne--ity arthaḥ
% madhyamaveśmani might denote 'in an inner/central chamber of the house.'
% pragharṣa = gharṣaṇa (Āyurvedīya Kośa), and gharṣaṇa = mardana. mardana = aṅgamardana
% udvartanaṃ = gātramardanaṃ (Āyurvedīya Kośa); 1. tvagavagharṣaṇa 2. snānāt pūrvaṃ tvagavagharṣaṇārtham upayujyamānaṃ cūrṇam āmalakyādi. 3. pādābhyām aṅgāvayavānāṃ pīḍanaṃ yena vyāyāmajanitaḥ śramaḥ śāmyati.

\item[14cd] % not in the vulgate
And if one is injured, one should avoid morbid conditions (\emph{doṣa}), such as moisture, and be protected by an umbrella.

\item[15ab]
One also should travel by means of female elephants and adorn oneself with the excellent aloe-wood.%
	\footnote{Ḍalhaṇa \citep[809]{vulgate} remarks that one should travel on female elephants to avoid mud and water, and one should wear aloe-wood because it wards off the cold (\emph{nāgavadhūbhir iti hastinīgamanaṃ ca kardamāmbunoḥ parihārārtham| praśastāgurubhūṣita iti aguruṇaḥ, śaityanivārakatvāt}).} %
Thus, [it has been taught].%
	\footnote{The \emph{iti} after 15ab appears to follow from verse 3. In K, the iti is followed by a rubricated section marker.}\\

%
% Then the vulgate has:
% divāsvapnam ajīrṇaṃ ca varjayet tatra yatnataḥ | 
% sevyāḥ śaradi yatnena kaṣāyasvādutiktakāḥ ||
% kṣīrekṣuvikṛtikṣaudraśālimudgādijāṅgalāḥ |
% śvetasrajaś candrapādāḥ pradoṣe laghu cāmbaram ||
% salilaṃ ca prasannatvāt sarvam eva tadā hitam
% saraḥsv āplavanaṃ caiva kamalotpalaśāliṣu ||

\item[15cd--17ab] % not in the vulgate
Homage to the sun, who has spread forth its light along the bright pathways of the quarters, when the sky's covering of dense blue is spread over the extent of the earth, and who is exalted by being born at the sight of Indra’s thunderbolt. The splendour of the sun advances, smothering cloud after cloud.
% I am assuming the locative is being used for the dative. Oberlies observes that namaskṛtya can take the locative in Epics (p 353), but also that the locative and dative are exchangeable. 
% not sure about °vaguṇṭhite vyomni. I've translated avaguṇṭhite more like a noun. 
%\skt{labdha-prasara}    mf(\skt{ā})n. that which has obtained free scope, moving at liberty, unimpeded Mudr. But does it  refer to the sun rising (i.e., when the sun has attained its rising) or the spreading forth of its light?

\item[17cd--18ab] % not in the vulgate
 Bile that is accumulated during the rains in autumn is aggravated by the heat. For this reason, when the clouds are dispersing, all people should undertake practices that reduce bile. 

\item[18cd--19ab] % not in the vulgate
During autumn, when one has an appetite, one should eat food that is cool, astringent and sweet, somewhat light and  bitter, as well as mixed with ghee and containing game.
% can food be both astringent and sweet or should we understand the compound kaṣāyamadhuram as astringent or sweet?
% I have read  īṣal in compound with laghu. 
% Is satiktakam bitter or pungent?
% jāṅgala appears to range from deer (jaṅghāla) to birds and burrowing animals (Ḍalhaṇa 1.46.53: jāṅgalaśabdena jaṅghālaviṣkirapratudaparṇamṛgabileśayā iti trayodaśa bhedāḥ ṣaṭsvevāntarbhūtāḥ)

\item[19cd--20ab] % not in the vulgate
If not indulging excessively in sex, one should enjoy eating deer, spotted antelopes, quails, goats, black antelopes, and  partridges, along with mung beans, aged rice, and barley.

\item[20cd--21ab] % not in the vulgate
One should avoid burning, acrid, spicy and heating foods, sleeping during the day, and heavy foods, as well as vegetables and meats that are heavy, and exposure to frost.

% It seems strange to have divāśvapna in the middle of a compound of types of food, so I suspect that at least one anusvāra has dropped out, more likely two (unless I've misunderstood gurūṇi. Note my emendations. The atha suggests at least two compounds were intended here (instead of the one in K and H).
% Cf. Aṣṭāṅgahṛdaya, nidānasthāna 13.2526
% vyādhikarmopavāsādikṣīṇasya bhajato drutam | 
% atimātram athānyasya gurvamlasnigdhaśītalam | 
% lavaṇakṣāratīkṣṇoṣṇaśākāmbu svapnajāgaram |
% mṛdgrāmyamāṃsavallūram ajīrṇaśramamaithunam ||

\item[21cd--22ab] % not in the vulgate
O King, in autumn, the wise person should carefully follow this rule:
One should consume grapes, sugarcane and milk in that season (\emph{tatra}).
% confirmation that a king is being spoken to here.

\item[22cd--23ab] 
When bile has accumulated during the rains, one should eliminate it with purgatives, or by administering bitter medicated ghee, and also by bloodletting from the veins.
% I assume that tiktaḥ and sarpis need to be read in compound with prayoga (and have emended tikta accordingly). 

\item[23cd--24] % In the vulgate marked by round brackets
Water should be tasty and cool, pure, clear like a flawless crystal, cleansed by the rays of the autumn sun, and detoxified by the rising star Agastya. Because of its clearness, all such water is then beneficial.

\item[25--26ab] %  In the vulgate marked by round brackets
And in autumn seasons, one should enjoy the breeze in lotus, lily and rice fields; or the sandalwood with fragrant camphor; clean light clothes; autumnal garlands; and within reason drinking sweet wine.
%
% pavanam ... kamalotpalaśāliṣu: the breeze blowing among lotuses, lilies and rice stalks?
% yuktitaḥ – within reason?

\item[26cd--27ab] % 26cd not in vulgate
One should do everything that pacifies bile: enjoying the moonlight, groups of friends, pleasant comforts, and sweet speech.
\end{translation}    
